A Road Segment Prioritization Approach for Cycling Infrastructure

### 2023-04-24 README.Rmd

geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates a new approach to support the prioritization of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis methods to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these methods has been applied to cycling infrastructure prioritization in previous research, this is the first time that they have been combined, creating an integrated road segment prioritization approach. The approach, which can be applied to other cities, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling networks, low-traffic neighborhoods, potential demand, routing"
output:
 # bookdown::github_document2:
   bookdown::pdf_document2:
 # bookdown::word_document2:
      highlight: null   # to remove code highlighting in r chunks
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
      # https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown
      # pandoc_args:
      # - '--lua-filter=extras/scholarly-metadata.lua'
      # - '--lua-filter=extras/author-info-blocks.lua'
      keep_tex:  true
editor_options:
  markdown:
    wrap: sentence
# elsevier style for referencing (from https://github.com/citation-style-language/styles)
csl: extras/elsevier-harvard.csl
always_allow_html: yes
bibliography: references.bib
bibliography: references.bib
---

@@

Other approaches have focused on characteristics of route segments as the basis of analysis to help prioritize investment in cycle infrastructure where it is most needed.
@grisé2018 developed a multi-criteria approach, combined with estimated levels of usage, to highlight routes most in need of investment in Quebec City.
@wysling2022 present a method that used open data from Paris to generate cycling suitability ratings based on 'bicycle suitability multipliers' associated with different types of road across the city, to support more effective interventions.

@@

<!-- EDIT! Add more to limitations -->

Our work builds on these community finding approaches by proposing a similar greedy network expansion algorithm for cycle network expansion *within communities*.
We incorporate community finding methods for study area partitioning with weighted routing to avoid links that are stressful to cycle on.
In doing so, we propose an approach that ensures a better distribution investments across the study area while also accounting for motivators and deterrents to cycling.
We propose three sub-methods that address some of the limitations of previous studies.
These limitations include (a) an inherent bias when basing network design solely on existing cycling demand, (b) the proposal of routes that may not correspond to studies on cyclist preference and government policies, and (c) an insufficient consideration of the ethical principles underlying the analysis.
The current work addresses these limitations through the following organisation.
Section \@ref(calculating-potential-cycling-demand) focuses on calculating potential cycling demand.
Section \@ref(routing) focuses on routing the demand onto the road network while accounting for cyclist preferences and government priorities.
Section \@ref(road-segment-prioritization) outlines a method for partitioning the study area based on a community finding algorithm and routed cycling demand.
It then introduces the network expansion algorithms, and compares an approach grounded in 'egalitarianism' to one grounded in 'utilitarianism'.

# Calculating Potential Cycling Demand

@@

*OD* data can be obtained from a range of sources, including GPS data, household travels surveys and census data on work locations.
In areas where observed OD data is unavailable, modelling techniques such as spatial interaction models can be used to estimate travel volumes between zones [@black1995spatial; @martinez2013new; @wilson1971family].
In this paper we use open access data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between administrative zones --- Middle layer Super Output Areas (MSOA) --- by mode of travel.
MSOAs have an average population of just over 8000 [@ofn2018population].[^3]
Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling) illustrate the proportion of trips cycled by distance and the geographic extent of the input OD dataset used in this paper.

@@

The above points are accounted for by using a weighted road network for routing.
This has been previously done by multiplying all road segments without cycling infrastructure by a fixed impedance factor [@mauttone2017bicycle], or by assigning different weights to the road segments proportional to the investment cost of bringing them to an acceptable level of stress for cycling [@duthie2014optimization].
Perceived stress for cyclists has been found to be a function of vehicular traffic volume and speed [@sorton1994bicycle], both of which vary predictably with road type.
@gehrke2020cycling build on this information to use routing impedance factors for road segments that are based on road type and the existence of cycling infrastructure.

For the purposes of this paper we created a similar weighting profile that is adjusted to favor less stressful roads (based on the definition by @sorton1994bicycle and information from Table \@ref(table:osmroadtypes)) and roads with existing cycling infrastructure.
We believe this to be more appropriate than the simplistic approach adopted by @mauttone2017bicycle as it makes use of a hierarchy of road preference based on perceived stress levels, going beyond simply favouring roads with existing cycling infrastructure.

>>>>

geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates a new approach to support the prioritization of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis methods to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these methods has been applied to cycling infrastructure prioritization in previous research, this is the first time that they have been combined, creating an integrated road segment prioritization approach. The approach, which can be applied to other cities, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling networks, low-traffic neighborhoods, potential demand, routing"
output:
 bookdown::github_document2:
   # bookdown::pdf_document2:
 # bookdown::word_document2:
      highlight: null   # to remove code highlighting in r chunks
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
      # https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown
      # pandoc_args:
      # - '--lua-filter=extras/scholarly-metadata.lua'
      # - '--lua-filter=extras/author-info-blocks.lua'
      keep_tex:  true
editor_options:
  markdown:
    wrap: sentence
# elsevier style for referencing (from https://github.com/citation-style-language/styles)
csl: extras/elsevier-harvard.csl
always_allow_html: yes
---

@@

Other approaches have focused on characteristics of route segments as the basis of analysis to help prioritize investment in cycle infrastructure where it is most needed.
@grise2018if developed a multi-criteria approach, combined with estimated levels of usage, to highlight routes most in need of investment in Quebec City.
@wysling2022improve present a method that used open data from Paris to generate cycling suitability ratings based on 'bicycle suitability multipliers' associated with different types of road across the city, to support more effective interventions.

@@

Our work builds on these community finding approaches by proposing a similar greedy network expansion algorithm for cycle network expansion *within communities*.
We incorporate community finding methods for study area partitioning with weighted routing to avoid links that are stressful to cycle on.
In doing so, we propose an approach that ensures a better distribution investments across the study area while also accounting for motivators and deterrents to cycling.
We propose three sub-methods that address some of the limitations of previous studies.
These limitations include (a) an inherent bias when basing network design solely on existing cycling demand, (b) the proposal of routes that may not correspond to studies on cyclist preference and government policies, and (c) an insufficient consideration of the ethical principles underlying the analysis.
The current work addresses these limitations through the following organisation.
Section \@ref(calculating-potential-cycling-demand) focuses on calculating potential cycling demand.
Section \@ref(routing) focuses on routing the demand onto the road network while accounting for cyclist preferences and government priorities.
Section \@ref(road-segment-prioritization) outlines a method for partitioning the study area based on a community finding algorithm and routed cycling demand.
It then introduces the network expansion algorithms, and compares an approach grounded in 'egalitarianism' to one grounded in 'utilitarianism'. ithms, and compares an approach grounded in
'egalitarianism' to one grounded in 'utilitarianism'. As mentioned above, Manchester is used as a case study. 
An overview of the steps involved is shown in Figure \@ref(fig:methodflowchart).

```{r methodflowchart, fig.cap="Flowchart of proposed methodology", out.width="80%", fig.show='hold', fig.align = "center", echo=FALSE}
knitr::include_graphics("extras/methodology_flowchart.png")
```

## Calculating Potential Cycling Demand

@@

*OD* data can be obtained from a range of sources, including GPS data, household travels surveys and census data on work locations.
In areas where observed OD data is unavailable, modelling techniques such as spatial interaction models can be used to estimate travel volumes between zones [@black1995spatial; @martinez2013new; @wilson1971family].
In this paper we use open access data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between administrative zones --- Middle layer Super Output Areas (MSOA) --- by mode of travel (this is similar to the approach used by @grise2018if).
MSOAs have an average population of just over 8000 [@ofn2018population].[^3]
Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling) illustrate the proportion of trips cycled by distance and the geographic extent of the input OD dataset used in this paper.

@@

The above points are accounted for by using a weighted road network for routing.
This has been previously done by multiplying all road segments without cycling infrastructure by a fixed impedance factor [@mauttone2017bicycle], or by assigning different weights to the road segments that are (a) proportional to the investment cost of bringing them to an acceptable level of stress for cycling [@duthie2014optimization], or (b) inversely proportional to perceived level of stress experienced by cyclists [@wysling2022improve].
Perceived stress for cyclists has been found to be a function of vehicular traffic volume and speed [@sorton1994bicycle], both of which vary predictably with road type.
@gehrke2020cycling build on this information to use routing impedance factors for road segments that are based on road type and the existence of cycling infrastructure.

For the purposes of this paper we created a similar weighting profile that is adjusted to favor less stressful roads (based on the definition by @sorton1994bicycle and information from Table \@ref(table:osmroadtypes)) and roads with existing cycling infrastructure.
We believe this to be more appropriate than the simplistic approach adopted by @mauttone2017bicycle as it makes use of a hierarchy of road preference based on perceived stress levels, going beyond simply favouring roads with existing cycling infrastructure.
It is also different to the weighting profile used by @wysling2022improve as it adds a larger weight on trunk and primary roads then other roads without cycling infrastructure.

### 2023-04-24 README.tex

The 2015 Paris agreement (UN, 2015) acknowledged that
fundamental changes to societies and economies are necessary to mitigate
climate change. Like other sectors, transport is under substantial
pressure to decarbonise, resulting in a number of technical innovations
including electric vehicles. But new vehicle technologies can only go so
far and do not tackle parallel problems such as congestion, road traffic
casualties and physical inactivity (Brand et al., 2020).

In this context, interest and investment in active modes are growing.
The benefits extend beyond congestion and the environment, as active
travel can also help alleviate what is referred to as the pandemic of
global inactivity; physical inactivity is on the rise and has become the
4th highest cause of death globally (Kohl 3rd et al., 2012). Various
studies have documented the association between active transport and
lower risk of disease, including cancer and cardiovascular disease
(Celis-Morales et al., 2017; Jarrett et al., 2012; Patterson et al., 2020). In the wake of the Covid-19 pandemic, and
the resulting reduced capacity of public transport, the UK government
has pledged to invest billions of pounds to improve walking and cycling
infrastructure across the country. While this unprecedented sum is an
opportunity to reshape cities in a way that improves the well-being of
citizens, it does come with a warning:

@@

The funding on its own is therefore no guarantee of a change in
commuting across the country; it must be used to design adequate cycling
infrastructure that is based on motivators and deterrents to cycling. In this paper we
propose an approach for planning a connected cycling network that is meant to
be applicable to any city (as shown in the Appendix), and we use the city of Manchester as a case study. Greater
Manchester aims to become the first region to reach a carbon neutral transport network;
its local authorities have developed plans for a 1800 mile walking and cycling network
(Bee Network) that will be delivered over 10 years (Greater Manchester, 2020). While schemes
have already been put in place for around a quarter of this network, the majority is yet to be built.

@@

Much research has been done to understand what gets people to cycle.
Segregated cycling infrastructure\footnote{\emph{Segregated cycling infrastructure}
  refers to road space that is allocated to cyclists only, with physical
  separation to protect cyclists from other modes of transport.} has been
shown to increase cycling uptake (Aldred et al., 2019; Goodman et al., 2014; Marqués et al., 2015), with the separation from motorized
vehicles being key (Winters et al., 2011). Revealed preference of
cyclists shows that they are willing to deviate from the most efficient
routes in order to commute on safer roads (Crane et al., 2017).
However, such deviations are only considered if they do not considerably
increase route circuitry; behaviour studies have found that the
probability of choosing a route decreases in proportion to its length
relative to the shortest route (Broach et al., 2011; Winters et al., 2010).
Another defining feature for cycling infrastructure is how well
connected it is. Cyclists prefer cohesive infrastructure, particularly
when cycling on arterial roads with high levels of motorized traffic
(Stinson and Bhat, 2003). The lack of well-connected cycling
infrastructure is one of the main obstacles to increasing cycling uptake
(Caulfield et al., 2012). While direct and cohesive cycling networks
have been shown to positively impact cycling rates, density\footnote{making an
  area's bicycle network denser means adding more cycling routes in the
  area and thereby giving cyclists more route options.} of the cycling
network is also vital (Schoner and Levinson, 2014).

@@

\emph{Optimization} methods have been used to propose improvements to
cycling networks. Mesbah et al. (2012) propose a bi-level formulation to
optimize allocation of cycling lanes to the network without exceeding a
set budget. They account for the effect of cycling lanes on car traffic,
and attempt to maximize utilization of said lanes with minimal impact on
car travel times. Safeguarding against increased car traffic may be counter-productive if the goal is to create a mode-shift, as research has shown that reducing road space for cars leads to less cars on the road, a phenomenon referred to as ``traffic evaporation'' (Nello-Deakin, 2020).
On the other hand, Mauttone et al. (2017) developed an optimization framework aiming at minimizing the total
user cost of cycling on the network without considering car usage. The aggregate flow\footnote{\emph{flow} is
  used throughout this research to refer to the cycling demand when it is
  routed onto the road network. The flow on any road segment is the
  cumulative demand on it, resulting from cyclists commuting between
  various Origin-Destination (OD) pairs.} on road segments (links) is obtained by using
shortest paths to route existing cycling demand onto the road network. The solution is a proposed set of links where cycling infrastructure
should be added in order to minimize the overall travel cost of cyclists
across the network. The cost of traversing a link is given as a function
of its length and whether or not it has cycling infrastructure. The
problem has also been solved by attempting to find the minimum cost of
improving roadway links to meet a desired level of service (LOS)
(Duthie and Unnikrishnan, 2014). In this formulation, all origin-destination
(OD) pairs need to be connected by roads that meet the desired LOS. A
limitation of these approaches is that they do not explicitly solve for
continuity. Continuity is addressed using either a constraint specifying
that each link with a bike lane should be connected to at least one
destination (Mesbah et al., 2012), a constraint on maximum deviation
from shortest paths (Duthie and Unnikrishnan, 2014), or a discontinuity
penalty to prioritize connected road segments (Mauttone et al., 2017).

In this paper, continuity is analysed by looking at the connectivity of
the network through the graph-theoretic concept of \emph{connected
components}. Orozco et al. (2020) study the existing cycling network in terms of its disconnected
components. They propose two different algorithms to connect disconnected components; one connects the two largest components, and the other connects the largest component to its nearest neighbor. By measuring the growth of the largest connected component as a function of the kilometers of network added, they determine that these approaches are more effective at improving cycling network connectivity than random allocation of cycling infrastructure.
The concept of connected components is
also at the core of the methodology proposed by Olmos et al. (2020). After
routing the cycling demand onto the network links, they use percolation
theory to filter out the links based on the aggregate flow passing through them, varying the flow threshold for
filtering to identify the minimum flow at which the whole city is
connected by a giant component. While these approaches deal with
continuity better, they look at the network as a whole when attempting
to improve it, and in doing so fail to account for equitable
distribution of infrastructure.

@@

Our work builds on these community finding approaches by proposing a
similar greedy network expansion algorithm for cycle network expansion
\emph{within communities}. We incorporate community finding methods for study
area partitioning with weighted routing to avoid links that are
stressful to cycle on. In doing so, we propose an approach that ensures a better distribution
investments across the study area while also accounting for motivators and deterrents to cycling.
We propose three sub-methods that address some of the limitations of previous
studies. These limitations include (a) an inherent bias when basing network
design solely on existing cycling demand, (b) the proposal of routes that may
not correspond to studies on cyclist preference and government policies,
and (c) an insufficient consideration of the ethical principles underlying
the analysis. The current work addresses these limitations through the following organisation. Section \ref{calculating-potential-cycling-demand}
focuses on calculating potential cycling demand. Section \ref{routing}
focuses on routing the demand onto the road network while accounting for
cyclist preferences and government priorities. Section
\ref{road-segment-prioritization} outlines a method for partitioning the
study area based on a community finding algorithm and routed cycling
demand. It then introduces the network expansion algorithms, and compares an approach grounded in
`egalitarianism' to one grounded in `utilitarianism'. As mentioned above, Manchester is used as a case study.
An overview of the steps involved is shown in Figure \ref{fig:methodflowchart}.

@@

\emph{OD} data can be obtained from a range of sources,
including GPS data, household travels surveys and census data on work
locations. In areas where observed OD data is unavailable, modelling
techniques such as spatial interaction models can be used to estimate
travel volumes between zones (Black, 1995; Martínez and Viegas, 2013; Wilson, 1971). In this paper we use open access data from the UK
census (ONS, 2011), which contains aggregate statistics on
number of commuters between administrative zones --- Middle layer Super
Output Areas (MSOA) --- by mode of travel (this is similar to the approach
used by Grisé and El-Geneidy (2018)).
MSOAs have an average
population of just over 8000 (ONS, 2018).\footnote{ See
  \url{https://wicid.ukdataservice.ac.uk/} for open access to the OD data.}
Figures \ref{fig:potdemhistograms} and \ref{fig:desirefacetcycling}
illustrate the proportion of trips cycled by distance and the
geographic extent of the input OD dataset used in this paper.

@@

The above points are accounted for by using a weighted road network for
routing. This has been previously done by multiplying all road segments
without cycling infrastructure by a fixed impedance factor
(Mauttone et al., 2017), by assigning different weights to the road
segments that are (a) proportional to the investment cost of bringing them to an
acceptable level of stress for cycling (Duthie and Unnikrishnan, 2014), or (b) inversely proportional to
perceived level of stress experienced by cyclists (Wysling and Purves, 2022).
Perceived stress for cyclists has been found to be a function of vehicular traffic
volume and speed (Sorton and Walsh, 1994), both of which vary predictably
with road type. Gehrke et al. (2020) build on this information to use
routing impedance factors for road segments that are based on road type
and the existence of cycling infrastructure.

>>>>

The 2015 Paris agreement (UN, 2015) acknowledged that fundamental changes to societies and economies are necessary to mitigate climate change.
Like other sectors, transport is under substantial pressure to decarbonise, resulting in a number of technical innovations including electric vehicles.
But new vehicle technologies can only go so far and fail to tackle parallel problems such as congestion, road traffic casualties and physical inactivity (Brand et al., 2020).

In this context, interest and investment in active modes has grown substantially, especially with the growing realisation that the benefits extend beyond congestion and the environment, as active travel can also help alleviate what is referred to as the pandemic of global inactivity; physical inactivity is on the rise and has become the 4th highest cause of death globally (Kohl 3rd et al., 2012).
Various studies have documented the association between active transport and lower risk of disease, including cancer and cardiovascular disease (Celis-Morales et al., 2017; Jarrett et al., 2012; Patterson et al., 2020).
In the wake of the Covid-19 pandemic, and the resulting reduced capacity of public transport, the UK government has pledged to invest billions of pounds to improve walking and cycling infrastructure across the country.
While this unprecedented sum is an opportunity to reshape cities in a way that improves the well-being of citizens, it does come with a warning:

@@

Funding alone is therefore no guarantee of a change in commuting across the country; it must be used to design adequate cycling infrastructure that is based on motivators and deterrents to cycling.
In this paper we propose an approach for planning connected cycling networks that can be generalised to any city with appropriate input data (as shown in the Appendix).
We used the city of Manchester as a case study due to its ambitions to become the first UK region with a carbon neutral transport network, supported by ambitious plans for a 1800 mile walking and cycling network (the `Bee Network') that will be delivered over 10 years (Greater Manchester, 2020).
While schemes have already been put in place for around a quarter of this network, the majority is yet to be built.

@@

Much research has explored the question of what gets people cycling.
Segregated cycling infrastructure --- road space that is allocated to cyclists only with physical separation to protect cyclists --- is associated with cycling uptake (Aldred et al., 2019; Goodman et al., 2014; Marqués et al., 2015).
Separation from motorized vehicles has found to be key (Winters et al., 2011).
Revealed preference of cyclists shows that they are willing to deviate from the most efficient routes in order to commute on safer roads (Crane et al., 2017).
However, such deviations are only considered if they do not considerably increase route circuitry; behavior studies have found that the probability of choosing a route decreases in proportion to its length relative to the shortest route (Broach et al., 2011; Winters et al., 2010).
Another defining feature for cycling infrastructure is how well connected it is.
Cyclists prefer cohesive infrastructure, particularly when cycling on arterial roads with high levels of motorized traffic (Stinson and Bhat, 2003).
The lack of well-connected cycling infrastructure is one of the main obstacles to increasing cycling uptake (Caulfield et al., 2012).
While direct and cohesive cycling networks have been shown to positively impact cycling rates, density\footnote{making an area's bicycle network denser means adding more cycling routes in the area and thereby giving cyclists more route options.} of the cycling network is also vital (Schoner and Levinson, 2014).

@@

\emph{Optimization} methods have been used to propose improvements to cycling networks.
Mesbah et al. (2012) propose a bi-level formulation to optimize allocation of cycling lanes to the network without exceeding a set budget.
They account for the effect of cycling lanes on car traffic, and attempt to maximize utilization of said lanes with minimal impact on car travel times.
Safeguarding against increased car traffic may be counter-productive if the goal is to create a mode-shift, as research has shown that reducing road space for cars leads to less cars on the road, a phenomenon referred to as ``traffic evaporation'' (Nello-Deakin, 2020).

Mauttone et al. (2017) developed an optimization framework aiming at minimizing the total user cost of cycling on the network without considering car usage.
The aggregate flow\footnote{\emph{flow} is used throughout this research to refer to the cycling demand when it is routed onto the road network.
  The flow on any road segment is the cumulative demand on it, resulting from cyclists commuting between various Origin-Destination (OD) pairs.} on road segments (links) is obtained by using shortest paths to route existing cycling demand onto the road network.
The solution is a proposed set of links where cycling infrastructure should be added in order to minimize the overall travel cost of cyclists across the network.
The cost of traversing a link is given as a function of its length and whether or not it has cycling infrastructure.
The problem has also been solved by attempting to find the minimum cost of improving roadway links to meet a desired level of service (LOS) (Duthie and Unnikrishnan, 2014).
In this formulation, all origin-destination (OD) pairs need to be connected by roads that meet the desired LOS.
A limitation of these approaches is that they do not explicitly solve for continuity.
Continuity is addressed using either a constraint specifying that each link with a bike lane should be connected to at least one destination (Mesbah et al., 2012), a constraint on maximum deviation from shortest paths (Duthie and Unnikrishnan, 2014), or a discontinuity penalty to prioritize connected road segments (Mauttone et al., 2017).
In this paper, continuity is analysed by looking at the connectivity of the network through the graph-theoretic concept of \emph{connected components}.

Orozco et al. (2020) study the existing cycling network in terms of its disconnected components.
They propose two different algorithms to connect disconnected components; one connects the two largest components, and the other connects the largest component to its nearest neighbor.
By measuring the growth of the largest connected component as a function of the kilometers of network added, they determine that these approaches are more effective at improving cycling network connectivity than random allocation of cycling infrastructure.
The concept of connected components is also at the core of the methodology proposed by Olmos et al. (2020).
After routing the cycling demand onto the network links, they use percolation theory to filter out the links based on the aggregate flow passing through them, varying the flow threshold for filtering to identify the minimum flow at which the whole city is connected by a single component.
While these approaches deal with continuity better, they look at the network as a whole when attempting to improve it, and in doing so fail to account for equitable distribution of infrastructure.

Other approaches have focused on characteristics of route segments as the basis of analysis to help prioritize investment in cycle infrastructure where it is most needed.
Grisé and El-Geneidy (2018) developed a multi-criteria approach, combined with estimated levels of usage, to highlight routes most in need of investment in Quebec City.
Wysling and Purves (2022) present a method that used open data from Paris to generate cycling suitability ratings based on `bicycle suitability multipliers' associated with different types of road across the city, to support more effective interventions.

None of the aforementioned studies explore the equity impacts of different network prioritization approaches.
The following section highlights the importance of considering (often unstated) ethical assumptions underlying network optimisation and prioritization approaches of the type outlined above.

@@

Our work builds on these community finding approaches by proposing a similar greedy network expansion algorithm for cycle network expansion \emph{within communities}.
We incorporate community finding methods for study area partitioning with weighted routing to avoid links that are stressful to cycle on.
In doing so, we propose an approach that ensures a better distribution investments across the study area while also accounting for motivators and deterrents to cycling.
We propose three sub-methods that address some of the limitations of previous studies.
These limitations include (a) an inherent bias when basing network design solely on existing cycling demand, (b) the proposal of routes that may not correspond to studies on cyclist preference and government policies, and (c) an insufficient consideration of the ethical principles underlying the analysis.
The current work addresses these limitations through the following organisation.
Section \ref{calculating-potential-cycling-demand} focuses on calculating potential cycling demand.
Section \ref{routing} focuses on routing the demand onto the road network while accounting for cyclist preferences and government priorities.
Section \ref{road-segment-prioritization} outlines a method for partitioning the study area based on a community finding algorithm and routed cycling demand.
It then introduces the network expansion algorithms, and compares an approach grounded in `egalitarianism' to one grounded in `utilitarianism'. ithms, and compares an approach grounded in
`egalitarianism' to one grounded in `utilitarianism'. As mentioned above, Manchester is used as a case study.
An overview of the steps involved is shown in Figure \ref{fig:methodflowchart}.

@@

\emph{OD} data can be obtained from a range of sources, including GPS data, household travels surveys and census data on work locations.
In areas where observed OD data is unavailable, modelling techniques such as spatial interaction models can be used to estimate travel volumes between zones (Black, 1995; Martínez and Viegas, 2013; Wilson, 1971).
In this paper we use open access data from the UK census (ONS, 2011), which contains aggregate statistics on number of commuters between administrative zones --- Middle layer Super Output Areas (MSOA) --- by mode of travel (this is similar to the approach used by Grisé and El-Geneidy (2018)).
MSOAs have an average population of just over 8000 (ONS, 2018).\footnote{See \url{https://wicid.ukdataservice.ac.uk/} for open access to the OD data.}
Figures \ref{fig:potdemhistograms} and \ref{fig:desirefacetcycling} illustrate the proportion of trips cycled by distance and the geographic extent of the input OD dataset used in this paper.

@@

The above points are accounted for by using a weighted road network for routing.
This has been previously done by multiplying all road segments without cycling infrastructure by a fixed impedance factor (Mauttone et al., 2017), or by assigning different weights to the road segments that are (a) proportional to the investment cost of bringing them to an acceptable level of stress for cycling (Duthie and Unnikrishnan, 2014), or (b) inversely proportional to perceived level of stress experienced by cyclists (Wysling and Purves, 2022).
Perceived stress for cyclists has been found to be a function of vehicular traffic volume and speed (Sorton and Walsh, 1994), both of which vary predictably with road type.
Gehrke et al. (2020) build on this information to use routing impedance factors for road segments that are based on road type and the existence of cycling infrastructure.
